HO unification from object language to meta language

### 2024-03-28 paper/main.tex
wip

\noindent
In both languages unification variables are identified by a natural number
representing a memory address.
The memory and its associated operations are described below:

@@

\begin{tabular}{ll}
  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
\end{tabular}
\vspace{5pt}

\noindent
When detailing examples we write links as equations between terms under a context.
The equality sign is subscripted with
kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A}{F_x~a}$ corresponds to:
\begin{elpicode}
abs x\ val (link-beta (uva A []) (app[uva F [x],con "a"]))
\end{elpicode}

@@

The first four rules are common to both equalities and correspond to $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

\noindent
The main point in showing these equality tests is to remark how weaker \Ee is,
and to identify the four rules that need special treatment in
the implementation of \Ue.

@@

\noindent
The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
variables). Elpi provides it as a builtin, but one could implement it by
systematically loading the hypothetical rule \elpiIn{name x} every time
a nominal constant is postulated via \elpiIn{pi x\ }.

@@

Applying the substitution corresponds to dereferencing a term with respect to
the memory. To ease the comparison we split \Fo dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, ans has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp},
in charge of ``flattening'' \elpiIn{fapp} nodes, has no corresponding
operation in \Ho. The reasons for this asymmetry is that an \elpiIn{fapp}
node with a flexible head is always mapped to a \elpiIn{uva} (as per
sections \ref{sec:compilation,sec:beta}), preventing nested applications to
materialize.

@@

\noindent
Note that the cut operator is inessential, it could be
removed at the cost of a verbose test on the head of \elpiIn{L}
in the last rule (\elpiIn{L} head
can be \elpiIn{fcon}, \elpiIn{flam} or a name).

Applying the substitution in \Ho{} is very similar, with
the caveat that assignments have to be moved to the
current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

>>>>

\noindent
Natural numbers represent the memory addresses that
identify unification variables in both languages.
The memory and its associated operations are described below:

@@

\begin{tabular}{ll}
  $f~ a$ &  \elpiIn{app[con "f", con "a"]}\\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} ~ a$ & \elpiIn{lam x\ app[uva F [x], con "a"]} \\
  $\lambda x.F_{x} ~ x$ & \elpiIn{lam x\ app[uva F [x], x]} \\
\end{tabular}
\vspace{5pt}

\noindent
When detailing examples we write links as equations between two
terms under a context.
The equality sign is subscripted with
kind of \elpiIn{baselink}. For example $\linkbetaM{x}{A_x}{F_x~a}$ corresponds to:
\begin{elpicode}
abs x\ val (link-beta (uva A [x]) (app[uva F [x],con "a"]))
\end{elpicode}

@@

The first four rules are common to both equalities
and just define the usual congruence over terms, and since
we use an HOAS encoding they also capture $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

\noindent
The main point in showing these equality tests is to remark how weaker \Ee is,
and to identify the four rules that need special treatment in
the implementation of \Uo.

@@

\noindent
The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
variables).\footnote{Elpi provides it as a builtin, but one could implement it by
systematically loading the hypothetical rule \elpiIn{name x} every time
a nominal constant is postulated via \elpiInFN{pi x\ }}
The choice of using n-ary application, rather than binary, is to make it
easy to access the application's head. The price we pay is that substituting
an application in the head of an application should be amended by
``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
we do in this presentation and, as we explain later,
no equivalent of \elpiIn{napp} is needed in \Ho.}
Finally note
Note that the cut operator is inessential, it could be
removed at the cost of a verbose test on the head of \elpiIn{L}
in the last rule (\elpiIn{L} head
can be \elpiIn{fcon}, \elpiIn{flam} or a name).

@@

Applying the substitution corresponds to dereferencing a term with respect to
the memory. To ease the comparison we split \Fo dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, and has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
has no corresponding operation in \Ho. The reasons for this asymmetry is
that an \elpiIn{fapp} node with a flexible head is always mapped
to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
preventing nested applications to materialize.

@@

\noindent
Applying the substitution in \Ho{} is very similar, with
the caveat that assignments have to be moved to the
current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.
